Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-09-20 paper-correlations.tex
Cleveref works bad with appendices
Prepare for submission

\bibliographystyle{elsarticle-num}
\journal{Computational Material Science}

@@

\author[Smith]{Vasily~Postnicov}
\author[Smith,MIT]{Aleksei~Samarin}
\author[Smith]{Marina ~V. Karsanina}
\author[Aus]{Mathieu ~Gravey}
\author[MIPT]{Aleksey~Khlyupin}
\author[Smith]{Kirill ~M. Gerke}
\ead{kg@ifz.ru}

@@

\title{Evaluation of classical correlation functions from 2/3D images on CPU and
  GPU architectures: introducing CorrelationFunctions.jl}

\begin{abstract}
Correlation functions are becoming one of the major tools for quantification of
structural information that is usually represented as 2D or 3D image of porous
material.  In this paper we introduce \code{CorrelationFunctions.jl} open-source
package developed in Julia and capable of computing all classical correlation
functions based on imaging input data. Images include both binary and
multi-phase representations.  Our code is capable of evaluating two-point
probability $S_2$, phase cross-correlation $\rho_{ij}$, cluster $C_2$,
lineal-path $L_2$, surface-surface $F_{ss}$, surface-void $F_{sv}$, pore-size
$P$ and chord-length $p$ distribution functions on both CPU and GPU
architectures.  Where possible, we presented two types of computations: full
correlation map (correlations of each point with other points on the image, that
also allows obtaining ensemble averaged CF) and directional correlation
functions (currently in major orthogonal and diagonal directions).  Such an
implementation allowed for the first time to assemble a completely free solution
to evaluate correlation functions under any operating system with well
documented application programming interface (API). Our package includes
automatic tests against analytical solutions that are described in the paper. We
measured execution times for all CPU and GPU implementations and as a rule of
thumb full correlation maps on GPU are faster than other methods. However, full
maps require more RAM and, thus, are limited to available RAM resources. On the
other hand, directional CFs are memory efficient and can be evaluated for huge
datasets -- this way they are the first candidates for structural data
compression of feature extraction. The package itself is available through Julia
package ecosystem and on GitHub
(\url{https://github.com/fatimp/CorrelationFunctions.jl}), the latter source
also contains documentation and additional helpful resources such as tutorials.
We believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} presented in this paper will significantly
facilitate the usage of correlation functions in numerous areas of structural
description and research of porous materials, as well as in machine learning
applications. We also present some example as applied to ceramic, soil composite
and oil-bearing rock samples based on their 3D X-ray tomography and 2D scanning
electron microscope images. Finally, we conclude our paper with discussion of
possible ways to further improve presented computational framework.
\end{abstract}

\maketitle

@@

\subsection{Full correlation map}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|^2] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of indicator function $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described in
\cref{sec:number-of-trials}. The following algorithm summarizes all the above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}|^2 / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation is $O(M \log M)$, where $M$ is the
number of elements in the input image.

@@

An algorithm for computation of linear-path function for one-dimensional slice
is presented below. Here $A$ is a one-dimensional input array, $phase$ can be
void, solid or some other phase, and $periodic$ is a boolean value which equals
to true if we compute $L_2$ in periodic mode and false otherwise. The efficiency
of the proposed algorithm is $O(n)$ where $n$ is the length of the array.
\begin{algorithmic}[1]
  \Procedure{$L_2$}{$A, phase, periodic$}
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \State $runs \gets countruns(A, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(result, run)$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(A)$
      \State $last \gets last(A)$
      \If{$first = last$}
        \State $updateperiodic(result, first, last)$
      \EndIf
    \EndIf
    \State \textbf{return} $success$ divided by number of trials
    (see \cref{sec:number-of-trials})
  \EndProcedure
  \\
  \Procedure{countruns}{$A, phase$}
    \State $runslist \gets [\ ]$
    \State $runs \gets 0$
    \ForAll{$x \in A$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \Comment Add $runs$ to the accumulator
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$A, runs$}
    \State $r \gets runs$
    \For{$idx \gets 0, runs - 1$}
      \State $A[idx] \gets A[idx] + r$
      \State $r \gets r - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$A, first, last$}
    \State $sum \gets first + last$
    \For{$idx \gets 0, sum - 1$}
      \State $up \gets \min(idx, first, last, sum - idx)$
      \State $A[idx] \gets A[idx] + up$
    \EndFor
  \EndProcedure
\end{algorithmic}

@@

\bibliography{apssamp}
\end{document}

>>>>

\journal{Computer Physics Communications}

@@

\begin{frontmatter}
  \title{Evaluation of classical correlation functions from 2/3D images on CPU and
    GPU architectures: introducing CorrelationFunctions.jl}

  \author[Smith]{Vasily~Postnicov}
  \author[Smith,MIT]{Aleksei~Samarin}
  \author[Smith]{Marina~V.~Karsanina}
  \author[Aus]{Mathieu~Gravey}
  \author[MIPT]{Aleksey~Khlyupin}
  \author[Smith]{Kirill~M.~Gerke\corref{boss}}
  \cortext[boss]{Corresponding author.\\\textit{E-mail address:} kg@ifz.ru}

@@

  \begin{abstract}
    Correlation functions are becoming one of the major tools for quantification
    of structural information that is usually represented as 2D or 3D images of
    porous material.  In this paper we introduce \code{CorrelationFunctions.jl}
    open-source package developed in Julia and capable of computing all
    classical correlation functions based on imaging input data. Images include
    both binary and multi-phase representations.  Our code is capable of
    evaluating two-point probability $S_2$, phase cross-correlation $\rho_{ij}$,
    cluster $C_2$, lineal-path $L_2$, surface-surface $F_{ss}$, surface-void
    $F_{sv}$, pore-size $P$ and chord-length $p$ distribution functions on both
    CPU and GPU architectures.  Where possible, we presented two types of
    computations: full correlation map (correlations of each point with other
    points on the image, that also allows obtaining ensemble averaged CF) and
    directional correlation functions (currently in major orthogonal and
    diagonal directions).  Such an implementation allowed for the first time to
    assemble a completely free solution to evaluate correlation functions under
    any operating system with well documented application programming interface
    (API). Our package includes automatic tests against analytical solutions
    that are described in the paper. We measured execution times for all CPU and
    GPU implementations and as a rule of thumb full correlation maps on GPU are
    faster than other methods. However, full maps require more RAM and, thus,
    are limited to available RAM resources. On the other hand, directional CFs
    are memory efficient and can be evaluated for huge datasets -- this way they
    are the first candidates for structural data compression of feature
    extraction. The package itself is available through Julia package ecosystem
    and on GitHub, the latter source also contains documentation and additional
    helpful resources such as tutorials.  We believe that a single powerful
    computational tool such as \code{CorrelationFunctions.jl} presented in this
    paper will significantly facilitate the usage of correlation functions in
    numerous areas of structural description and research of porous materials,
    as well as in machine learning applications. We also present some example as
    applied to ceramic, soil composite and oil-bearing rock samples based on
    their 3D X-ray tomography and 2D scanning electron microscope
    images. Finally, we conclude our paper with discussion of possible ways to
    further improve presented computational framework.
\end{abstract}

  \begin{keyword}
    3D images; correlation functions; image analysis
  \end{keyword}
\end{frontmatter}

{\bf PROGRAM SUMMARY} \\
\begin{small}
\noindent
{\em Program Title:} CorrelationFunctions.jl \\
{\em CPC Library link to program files:} (to be added by Technical Editor) \\
{\em Developer's repository link:}
\url{https://github.com/fatimp/CorrelationFunctions.jl} \\
{\em Code Ocean capsule:} (to be added by Technical Editor)\\
{\em Licensing provisions:} MIT \\
{\em Programming language:} Julia \\
{\em Supplementary material:} \\
{\em Nature of problem:} Correlation functions are invaluable universal
statistical descriptors of structure of porous media used in many scientific
field such as astronomy, soil physics, hydrology and biology. We propose an open
source and cross-platform solution for computation of vast majority of
correlation functions for both 2D and 3D images of porous material. \\
{\em Solution method:} Our package computes correlation functions using fast
Fourier transform using such preprocessing steps as cluster labeling or edge
detection if needed. Almost all of proposed algorithms can be executed on both
CPU and GPU which results in high execution speed on modern hardware. Two
approaches to evaluation of correlation functions are proposed: computation of
correlation maps and computation along predefined directions. These two
approaches can be thought of as an execution time - memory trade-off. \\
\end{small}

@@

\subsection{Full correlation map}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|^2] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of indicator function $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described in
\ref{sec:number-of-trials}. The following algorithm summarizes all the above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}|^2 / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation is $O(M \log M)$, where $M$ is the
number of elements in the input image.

@@

An algorithm for computation of linear-path function for one-dimensional slice
is presented below. Here $A$ is a one-dimensional input array, $phase$ can be
void, solid or some other phase, and $periodic$ is a boolean value which equals
to true if we compute $L_2$ in periodic mode and false otherwise. The efficiency
of the proposed algorithm is $O(n)$ where $n$ is the length of the array.
\begin{algorithmic}[1]
  \Procedure{$L_2$}{$A, phase, periodic$}
    \State $len \gets length(A)$
    \State $result \gets zeros(len)$
    \State $runs \gets countruns(A, phase)$
    \ForAll{$run \in runs$}
      \State $updateruns(result, run)$
    \EndFor
    \If{$periodic$}
      \State $first \gets first(A)$
      \State $last \gets last(A)$
      \If{$first = last$}
        \State $updateperiodic(result, first, last)$
      \EndIf
    \EndIf
    \State \textbf{return} $success$ divided by number of trials
    (see \ref{sec:number-of-trials})
  \EndProcedure
  \\
  \Procedure{countruns}{$A, phase$}
    \State $runslist \gets [\ ]$
    \State $runs \gets 0$
    \ForAll{$x \in A$}
      \If{$x = phase$}
        \State $runs \gets runs + 1$
      \ElsIf{$runs \ne 0$}
        \Comment Add $runs$ to the accumulator
        \State $runslist \gets runs:runslist$
        \State $runs \gets 0$
      \EndIf
    \EndFor
    \If{$runs \ne 0$}
      \State $runslist \gets runs:runslist$
    \EndIf
    \State \textbf{return} $runslist$
  \EndProcedure
  \\
  \Procedure{updateruns}{$A, runs$}
    \State $r \gets runs$
    \For{$idx \gets 0, runs - 1$}
      \State $A[idx] \gets A[idx] + r$
      \State $r \gets r - 1$
    \EndFor
  \EndProcedure
  \\
  \Procedure{updateperiodic}{$A, first, last$}
    \State $sum \gets first + last$
    \For{$idx \gets 0, sum - 1$}
      \State $up \gets \min(idx, first, last, sum - idx)$
      \State $A[idx] \gets A[idx] + up$
    \EndFor
  \EndProcedure
\end{algorithmic}

@@

\bibliographystyle{elsarticle-num}
\bibliography{apssamp}
\end{document}
